---
title: "Getting Started with the ODP SDK"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{odp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Ocean Data Platform (ODP) R SDK provides lightweight helpers for working
with tabular datasets exposed through the HubOcean API. This vignette walks you
through installing the package, authenticating, and streaming a dataset into R
structures that you probably already use.

## Installation

Install straight from GitHub with either `remotes`, `pak`, or `devtools`. The
examples below assume `remotes`:

```{r install, eval = FALSE}
install.packages("remotes") # once per machine
remotes::install_github("C4IROcean/odp-sdkr")
```

Load the package once it is installed:

```{r load, eval = FALSE}
library(odp)
```

## Authentication

Create a client with an API key. Either pass the key explicitly or set the
`ODP_API_KEY` environment variable before instantiating the client.

```{r auth, eval = FALSE}
client <- odp_client(api_key = "Sk_live_your_key")

# or keep the key out of scripts
Sys.setenv(ODP_API_KEY = "Sk_live_your_key")
client <- odp_client()
```

## Working with a Dataset

Use the dataset identifier you normally copy from the HubOcean catalog
(<https://app.hubocean.earth/catalog>). This vignette references the public
GLODAP dataset, but you can substitute your own ID.

```{r dataset, eval = FALSE}
dataset <- client$dataset("aea06582-fc49-4995-a9a8-2f31fcc65424")
table <- dataset$table
```

Inspect the schema and stats to plan queries:

```{r meta, eval = FALSE}
schema <- table$schema()
stats <- table$stats()
```

## Streaming Rows

`table$select()` returns an `OdpCursor` that lazily streams Arrow RecordBatches.
Iterate over the batches or materialise the results when you are ready.

```{r select, eval = FALSE}
cursor <- table$select(
  filter = "depth > $min_depth",
  vars = list(min_depth = 300),
  columns = c("latitude", "longitude", "depth"),
  timeout = 15
)

while (!is.null(batch <- cursor$next_batch())) {
  cat("chunk rows:", batch$num_rows, "\n")
}

df <- cursor$dataframe()
arrow_tbl <- cursor$arrow()
# tibble support is optional
# tib_tbl <- cursor$tibble()
```

Materialisation helpers (`dataframe()`, `arrow()`, `tibble()`) only drain the
batches that have not been streamed yet. Create a new cursor if you iterate with
`next_batch()` and later need the entire result set.

## Aggregations

Let the backend do the heavy lifting when you only need grouped stats.

```{r agg, eval = FALSE}
agg <- table$aggregate(
  group_by = "G2year",
  filter = "G2year >= 2020 AND G2year < 2025",
  aggr = list(G2salinity = "mean", G2tco2 = "max")
)
print(agg)
```

Each `aggr` entry specifies which reducer to apply (`"sum"`, `"min"`, `"max"`,
`"count"`, `"mean"`). You can also pass expressions (for example
`h3(geometry, 6)`) through `group_by`.

## Troubleshooting

- Install the `arrow` package; IPC streams power the transport layer
- Increase `timeout` for very large scans
- Keep your `ODP_API_KEY` secret—never commit it to version control

You can now explore ODP datasets from scripts, notebooks, or Shiny dashboards in
plain R.
