---
title: "Working with ODP Tabular Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{odp-tabular}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most ODP workflows revolve around tabular resources exposed through the HubOcean
API. This vignette focuses on those helpers: inspecting schemas, streaming data
in batches, projecting and filtering columns, letting the backend aggregate
rows for you, and modifying data (inserting, replacing, or deleting rows).

## Installation

Install the development build straight from GitHub with `remotes`, `pak`, or
`devtools`. These examples use `remotes`:

```{r install, eval = FALSE}
install.packages("remotes") # once per machine
remotes::install_github("C4IROcean/odp-sdkr")

# local checkout? build vignettes so this article is included
# remotes::install_local("~/dev/odp_sdkr", build = TRUE, build_vignettes = TRUE)
```

```{r load, eval = FALSE}
library(odp)
```

Create a client with `odp_client()`, passing the API key explicitly or via the
`ODP_API_KEY` environment variable:

```{r auth, eval = FALSE}
client <- odp_client(api_key = "Sk_live_your_key")

Sys.setenv(ODP_API_KEY = "Sk_live_your_key")
client <- odp_client()
```

## Locate a tabular dataset

Dataset identifiers match the UUIDs in the HubOcean catalog
(<https://app.hubocean.earth/catalog>). This vignette references the public
GLODAP dataset, but any tabular dataset works.

```{r dataset, eval = FALSE}
dataset <- client$dataset("aea06582-fc49-4995-a9a8-2f31fcc65424")
table <- dataset$table
```

Use `table$schema()` and `table$stats()` to understand the shape of the table
before streaming rows:

```{r meta, eval = FALSE}
schema <- table$schema()
stats <- table$stats()
```

## Streaming batches

`table$select()` returns an `OdpCursor` that lazily streams Arrow
`RecordBatch` chunks. Iterate with `next_batch()` or materialise unread batches
into familiar R structures.

```{r select, eval = FALSE}
cursor <- table$select()

# Fetch the next batch in the format you prefer
batch <- cursor$next_batch() # as Arrow RecordBatch
df <- cursor$next_dataframe() # as base data.frame

# Or iterate over the whole table
while (!is.null(batch <- cursor$next_batch())) {
  cat("chunk rows:", batch$num_rows, "\n")
}

# Or fetch everything at once
df <- cursor$all_dataframe()
arrow_tbl <- cursor$all_table()
# optional tidyverse helper
# tib     <- cursor$all_tibble()
```

`all_dataframe()`/`all_table()`/`all_tibble()` accept `max_rows` and `max_time`
safety limits (defaults: 1 000 000 rows, 60 s). Use `next_batch()` for streaming
without limits.

### Filters

Filters accept SQL/Arrow-style expressions, including geospatial helpers like
`within`, `contains`, and `intersect`.

```{r filters, eval = FALSE}
cursor <- table$select(filter = "G2year >= 2020 AND G2year < 2025")

bbox <- 'geometry within "POLYGON ((-10 50, -5 50, -5 55, -10 55, -10 50))"'
cursor <- table$select(filter = bbox)
```

### Column projections

Restrict the projection when you only need specific fields:

```{r columns, eval = FALSE}
cursor <- table$select(columns = c("G2tco2", "G2year"))
```

### Bind variables

Bind parameters inside the filter to avoid string pasting. Use named or
positional variables:

```{r vars, eval = FALSE}
cursor <- table$select(
  filter = "G2year >= $start AND G2year < $end",
  vars = list(start = 2020, end = 2025)
)

# positional form
table$select(filter = "G2year >= ? AND G2year < ?", vars = list(2020, 2025))
```

## Aggregations

`table$aggregate()` lets the backend perform grouped computations so you only
transfer summary rows.

```{r agg, eval = FALSE}
agg <- table$aggregate(
  group_by = "G2year",
  filter = "G2year >= 2020 AND G2year < 2025",
  aggr = list(G2salinity = "avg", G2tco2 = "max")
)
print(agg)
```

`aggr` entries specify which aggregation to apply (`"sum"`, `"min"`, `"max"`,
`"count"`, or `"mean"`). Expressions such as `h3(geometry, 6)` or
`bucket(depth, 0, 200, 400)` are accepted in `group_by` to build grid-based
reports.

## Writing data

Use `table$insert()` to add data to a table. Data is validated against the
table schema (required fields, types, nullability) and automatically committed:

```{r insert, eval = FALSE}
df <- data.frame(
  latitude = c(10.5, 20.3, 30.1),
  longitude = c(-50.2, -40.8, -30.5),
  depth = c(100L, 200L, 150L),
  stringsAsFactors = FALSE
)

table$insert(df)
```

For multi-step workflows (insert, replace, delete in a single transaction),
start a transaction explicitly:

```{r txn, eval = FALSE}
tx <- table$begin()
tx$insert(df)
tx$replace(query = "depth < 50", data = new_df)
tx$delete(query = "depth > 50")
tx$commit()
```

## Managing schema

Create a new table from a schema or directly from data:

```{r create, eval = FALSE}
schema <- arrow::schema(
  latitude = arrow::float64(),
  longitude = arrow::float64(),
  depth = arrow::int32()
)
table$create(schema)
# or from data: table$create(df)
```

Alter the schema and re-ingest data:

```{r alter, eval = FALSE}
new_schema <- arrow::schema(latitude = arrow::float64(), longitude = arrow::float64(), depth = arrow::float64())
table$alter(new_schema)
```

Remove all rows while preserving schema, or drop the table entirely:

```{r cleanup, eval = FALSE}
table$truncate() # remove rows
table$drop() # remove table
```

The SDK buffers rows and auto-flushes at 10,000 rows or 10 MB.

## Troubleshooting

- Install `arrow`; the SDK streams IPC batches
- Increase `timeout` for very large scans
- Keep API keys secret—never commit them to version control

These pieces let you stitch together pipelines, dashboards, or notebooks that
work with ODP tabular datasets entirely from R.
