---
title: "Working with ODP Tabular Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{odp-tabular}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most ODP workflows revolve around tabular resources exposed through the HubOcean
API. This vignette focuses on those helpers: inspecting schemas, streaming data
in batches, projecting and filtering columns, and letting the backend aggregate
rows for you.

> Scope: writing/mutating tables is still out-of-scope for the SDK. The helpers
> here are read-only and revolve around authenticated scans of Arrow batches.

## Installation

Install the development build straight from GitHub with `remotes`, `pak`, or
`devtools`. These examples use `remotes`:

```{r install, eval = FALSE}
install.packages("remotes")  # once per machine
remotes::install_github("C4IROcean/odp-sdkr")

# local checkout? build vignettes so this article is included
# remotes::install_local("~/dev/odp_sdkr", build = TRUE, build_vignettes = TRUE)
```

```{r load, eval = FALSE}
library(odp)
```

Create a client with `odp_client()`, passing the API key explicitly or via the
`ODP_API_KEY` environment variable:

```{r auth, eval = FALSE}
client <- odp_client(api_key = "Sk_live_your_key")

Sys.setenv(ODP_API_KEY = "Sk_live_your_key")
client <- odp_client()
```

## Locate a tabular dataset

Dataset identifiers match the UUIDs in the HubOcean catalog
(<https://app.hubocean.earth/catalog>). This vignette references the public
GLODAP dataset, but any tabular dataset works.

```{r dataset, eval = FALSE}
dataset <- client$dataset("aea06582-fc49-4995-a9a8-2f31fcc65424")
table <- dataset$table
```

Use `table$schema()` and `table$stats()` to understand the shape of the table
before streaming rows:

```{r meta, eval = FALSE}
schema <- table$schema()
stats <- table$stats()
```

## Streaming batches

`table$select()` returns an `OdpCursor` that lazily streams Arrow
`RecordBatch` chunks. Iterate with `next_batch()` or materialise unread batches
into familiar R structures.

```{r select, eval = FALSE}
cursor <- table$select()
while (!is.null(batch <- cursor$next_batch())) {
  cat("chunk rows:", batch$num_rows, "\n")
  # process each RecordBatch on the fly
}

df <- cursor$dataframe()
arrow_tbl <- cursor$arrow()
# optional tidyverse helper
tib_tbl <- cursor$tibble()
```

Materialisation helpers only drain batches that have not been streamed yet. To
collect the full result after iterating with `next_batch()`, start a new cursor
and call `dataframe()`/`arrow()`/`tibble()` before consuming chunks.

### Filters

Filters accept SQL/Arrow-style expressions, including geospatial helpers like
`within`, `contains`, and `intersect`.

```{r filters, eval = FALSE}
cursor <- table$select(filter = "G2year >= 2020 AND G2year < 2025")

bbox <- 'geometry within "POLYGON ((-10 50, -5 50, -5 55, -10 55, -10 50))"'
cursor <- table$select(filter = bbox)
```

### Column projections

Restrict the projection when you only need specific fields:

```{r columns, eval = FALSE}
cursor <- table$select(columns = c("G2tco2", "G2year"))
```

### Bind variables

Bind parameters inside the filter to avoid string pasting. Use named or
positional variables:

```{r vars, eval = FALSE}
cursor <- table$select(
  filter = "G2year >= $start AND G2year < $end",
  vars = list(start = 2020, end = 2025)
)

# positional form
table$select(filter = "G2year >= ? AND G2year < ?", vars = list(2020, 2025))
```

## Aggregations

`table$aggregate()` lets the backend perform grouped computations so you only
transfer summary rows.

```{r agg, eval = FALSE}
agg <- table$aggregate(
  group_by = "G2year",
  filter = "G2year >= 2020 AND G2year < 2025",
  aggr = list(G2salinity = "mean", G2tco2 = "max")
)
print(agg)
```

`aggr` entries specify which aggregation to apply (`"sum"`, `"min"`, `"max"`,
`"count"`, or `"mean"`). Expressions such as `h3(geometry, 6)` or
`bucket(depth, 0, 200, 400)` are accepted in `group_by` to build grid-based
reports.

## Troubleshooting

- Install `arrow`; the SDK streams IPC batches
- Increase `timeout` for very large scans
- Keep API keys secret—never commit them to version control

These pieces let you stitch together pipelines, dashboards, or notebooks that
work with ODP tabular datasets entirely from R.
